---
title: "quast_genome_assembly_QC"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(tidyverse)
```

Read QUAST results
```{r}
quast_stats <- read.delim("../output/genome_assembly_QC/quast_assembly_stats.tsv") %>%
  janitor::clean_names()
```

Calculate mean, maximum, and minimum of number of scaffolds, total length of scaffolds, N50, GC content
```{r}
mean <- quast_stats %>%
  select(x_contigs, total_length, n50, gc) %>%
  summarise(number_scaffolds = mean(x_contigs),
            total_length = mean(total_length),
            N50 = mean(n50),
            GC_perc = mean(gc))
mean
```

Check outliers: number of scaffolds
```{r}
boxplot(quast_stats$x_contigs)

filter(quast_stats, x_contigs > 280)
```

Result: 5 samples (19_1_9, 19_4_7, 19_6_2, 7_2_1, as5_2_4) have relatively large number of scaffolds.

Check outliers: total length
```{r}
boxplot(quast_stats$total_length)

filter(quast_stats, total_length > 8000000)
```

Result: 2 samples (19_1_9, 19_4_7) have relatively large genome sizes. 

Check outliers: GC content
```{r}
boxplot(quast_stats$gc)

filter(quast_stats, gc < 58)
# 3 samples (19_1_9, 19_4_7, as5_2_4) have relatively low GC content.
```

Result: 3 samples (19_1_9, 19_4_7, as5_2_4) have relatively low GC content.

```{r}
sessionInfo()
```

